---
title: "CCA Example"
output: html_document
---

## Data: ADT meta-analysis 

Here, we read in the data:
```{r, message = F}
library(tidyverse)

adt <- readRDS("../data/adt_data.RDS")
```

A useful way to subset the data is to retain only one effect size per contrast. Here, we choose the *last* contrast reported as measured by the timing post-intervention.
```{r}
# Subset to get last follow up time for each contrast
adt_example <- adt %>%
  group_by(studyid, groupid1, groupid2) %>%
  arrange(desc(estimingpost)) %>%
  slice(1) %>%
  ungroup()
```

To simplify things, we generate a few categorical variables:

- `g1out`: whether treatments were outpatient
- `multidrug`: whether patients presented using multiple drugs
- `long`: if the treatment was longer than 100 days
- `gXhi`: if group X had high intensity treatment (>1.5 hrs per week). **NOTE: I'd recommend using these for our example**.
```{r}
# Mutate to get easier names
dd <- adt_example %>%
  mutate(g1out = (g1loc == "2. Outpatient"), 
         multidrug = (dvmicro == "7. Mixed drugs"), 
         long = (g2txdays > 100)) %>%
  select(studyid, es_g, se_g, 
         g1hrs = g1hrsperweek, g2hrs = g2hrsperweek, 
         g1out, multidrug, long) %>%
  mutate(v_g = se_g^2, 
         g1hi = (g1hrs > 1.5), 
         g2hi = (g2hrs > 1.5))
```

# Models for example

```{r, echo = F}
n_study <- dd %>% distinct(studyid) %>% nrow()
n_effect <- dd %>% nrow()
```


It would seem that we should demonstrate both a complete- and available-case analysis.
Consider the subset of the ADT data that contains `r n_effect` effect estimates from `r n_study` studies. 
These effect estimates involve contrasts between groups in a study that are subjected to different treatment conditions, denoted in the data as *Group 1* and *Group 2*. 
Treatment conditions vary across studies, and include assertive continuing care (ACC), behavioral therapy, (CBT), motivational enhancement therapy (MET), and family therapy, as well as "practice as usual" control groups where subjects were referred to community services, but did not receive specific substance abuse treatment.

The researchers who collected these data extracted several covariates regarding the treatment type, intensity (in hours per week), and duration (in days), as well as information regarding the racial composition of each group in a contrast.
To demonstrate a complete- and available-case analysis, consider two covariates regarding whether each group received a high-intensity intervention (>1.5 hours per week).
Let $X_{i1} = 1$ if group 1 received a high-intensity intervention and $X_{i1} = 0$ if group 1 received a low-intensity intervention. 
Similarly, let $X_{i2} = 1$ if group 2 received a high intensity and $X_{i2} = 0$ if group 2 received a low-intensity intervention.

In the data, the treatment intensity is missing for some of the effects.
The table below shows that only 37 of the `r n_effect` have a reported treatment intensity for both groups.

```{r, echo = F}
mipat <- mice::md.pattern(dd %>% select(g1hi, g2hi))
counts <- as.integer(dimnames(mipat)[[1]][1:4])
mitab <- tibble(g1_intensity = rep(c("Observed", "Missing"), each = 2), 
                g2_intensity = rep(c("Observed", "Missing"), 2), 
                counts = counts) %>%
  mutate(percent = counts/n_effect)

knitr::kable(mitab, digits = 2)
```

For reference, the correlation between $X_{i1}$ and $X_{i2}$ among the effects where both are observed is `r cor(dd$g1hi, dd$g2hi, use = "pairwise.complete")`

A complete case analysis would include only the 37 effects for which both covariates are observed. 
Because studies produce multiple effect sizes, we use robust variance estimation to compute standard errors:

```{r}
# RVE
library(robumeta)

rv_mod_hiint <- robu(formula = es_g ~ g1hi + g2hi, 
                     data = dd, 
                     studynum = studyid, 
                     var.eff.size = v_g,
                     # rho = 0.5,
                     small = TRUE)
rv_mod_hiint
```

Note that by default, most programs that run meta-regressions will only use complete cases.

However, in the table above, we see that there are an additional 17 effects for which $X_{i1}$ is observed, but not $X_{i2}$.
Thus, one may be tempted to include those observations, but exclude $X_{i2}$ from the model. 
This is referred to as a shifting-case analysis or a shifting units of analysis.
```{r}
rv_mod_hiint_g1 <- robu(formula = es_g ~ g1hi, 
                        data = dd, 
                        studynum = studyid,
                        var.eff.size = v_g, 
                        small = TRUE)

rv_mod_hiint_g1
```
Note that this estimate is much smaller, by over one third, than the complete-case analysis above.

A similar model can be fit for $X_{i2}$, which yields a point estimate that is quite different (and even in a different direction) than the estimate in the complete-case model:
```{r}
rv_mod_hiint_g2 <- robu(formula = es_g ~ g2hi, 
                        data = dd, 
                        studynum = studyid,
                        var.eff.size = v_g, 
                        small = TRUE)

rv_mod_hiint_g2
```

## Correction for missingness

```{r, echo = F, include = F}
dd_r <- dd %>% # select just the relevant variables
  select(studyid, es_g, v_g, g1hi, g2hi)

# generate imputations
m <- 1000 # number of imputations
library(mice); set.seed(10)
dd_mi <- mice(dd_r, # impute missing Xs 
              m = m, 
              method = "midastouch") # Completely random Xs

# extract imputed data
com <- complete(dd_mi, "long", include = FALSE) %>%
  mutate(g1hi_na = rep(is.na(dd$g1hi), m), 
         g2hi_na = rep(is.na(dd$g2hi), m))

# summarize imputations
emp_bias_parmas <- com %>% 
  group_by(.imp) %>%
  summarize(H0_g1 = sum(g1hi == 0 & g1hi_na)/sum(g1hi_na == 0), # H0 for X1
            H1_g1 = sum(g1hi == 1 & g1hi_na)/sum(g1hi_na == 1), # H1 for X1
            H0_g2 = sum(g1hi == 0 & g1hi_na)/sum(g1hi_na == 0), # H0 for X2
            H1_g2 = sum(g1hi == 1 & g1hi_na)/sum(g1hi_na == 1), # H1 for X2
            p01 = sum(g2hi == 1 & g1hi == 0)/sum(g1hi == 0), # P[X2 = 1 | X1 = 0]
            p11 = sum(g2hi == 1 & g1hi == 1)/sum(g1hi == 1), # P[X2 = 1 | X1 = 1]
            v0_g1 = sum(v_g * (g1hi == 0))/sum(g1hi == 0), # mean var for X1 = 0
            v1_g1 = sum(v_g * (g1hi == 1))/sum(g1hi == 1), # mean var for X1 = 1
            v0_g2 = sum(v_g * (g2hi == 0))/sum(g2hi == 0), # mean var for X2 = 0
            v1_g2 = sum(v_g * (g2hi == 1))/sum(g2hi == 1), # mean var for X2 = 1
            psi1_g1 = glm(I(1-g1hi_na) ~ g1hi + es_g, family = "binomial")$coefficients[["es_g"]], # psi1 for X1
            psi1_g2 = glm(I(1-g2hi_na) ~ g2hi + es_g, family = "binomial")$coefficients[["es_g"]]) # psi1 for X2

# get params from fitted model
betas <- lapply(1:m, 
                FUN = function(i){
                  dat <- com %>% filter(.imp == i)
                  fit <- robu(es_g ~ g1hi + g2hi, data = dat, studynum = studyid, var.eff.size = v_g) # fit MR model
                  mod <- fit$reg_table # extract coefficients
                  beta1 <- mod[[2, 2]]
                  beta2 <- mod[[3, 2]]
                  tau2 = as.numeric(sensitivity(fit)$`Rho = 0.2`[7]) # extract variance components
                  return(tibble(beta1 = beta1, beta2 = beta2, tau2 = tau2))
                }) %>% 
  bind_rows()

# aggregate components for bias and compute bias for example
ebp <- bind_cols(emp_bias_parmas, betas) %>%
  mutate(
    bias_b0CC = H0_g1 * (v0_g1 + tau2) * psi1_g1, 
    bias_b1CC = (H1_g1 * (v1_g1 + tau2) - H0_g1 * (v0_g1 * tau2)) * psi1_g1, 
    omv_bias_b0 = beta2 * p01, 
    omv_bias_b1 = beta2 * (p11 - p01)
    )

write_csv(x = ebp, file = "../data/cca_bias_table.csv")
```

```{r}
summary(ebp$bias_b0CC)
summary(ebp$bias_b1CC)
summary(ebp$omv_bias_b0)
summary(ebp$omv_bias_b1)
```


# Other models fit

Potential models fit with RVE:
```{r}
# Mod 1a: Were interventions outpatient (mostly study-level)
rv_mod_outpatient <- robu(formula = es_g ~ g1out, 
                          data = dd, 
                          studynum = studyid, 
                          var.eff.size = v_g, 
                          small = TRUE)

# Mod 1b: Were interventions long (>100 days) (mostly study-level)
rv_mod_long <- robu(formula = es_g ~ long, 
                    data = dd, 
                    studynum = studyid, 
                    var.eff.size = v_g, 
                    small = TRUE)

# Mod 2a: Hours per week for both groups (complete cases)
rv_mod_hrs <-  robu(formula = es_g ~ g1hrs + g2hrs, 
                    data = dd,
                    studynum = studyid, 
                    var.eff.size = v_g,
                    # rho = 0.5,
                    small = TRUE)

# Mod 2b: Hi-intensity for each group (>1.5 hrs per week) complete cases
rv_mod_hiint <- robu(formula = es_g ~ g1hi + g2hi, 
                     data = dd, 
                     studynum = studyid, 
                     var.eff.size = v_g,
                     # rho = 0.5,
                     small = TRUE)

# Mod 2c: Hi-intensity for only group 1 (>1.5 hrs per week) shifting cases
rvmod_hiint_g1 <- robu(formula = es_g ~ g1hi, 
                       data = dd, 
                       studynum = studyid, 
                       var.eff.size = v_g,
                       # rho = 0.5,
                       small = TRUE)

rvmod_hiint_g2 <- robu(formula = es_g ~ g2hi, 
                       data = dd %>% mutate(g1hi = g1hrs > 1.5, 
                                            g2hi = g2hrs > 1.5), 
                       studynum = studyid, 
                       var.eff.size = v_g,
                       # rho = 0.5,
                       small = TRUE)
```


Potential parameters of interest
```{r}
mis_mech_df <- dd %>%
  mutate(long_na = is.na(long), 
         multidrug_na = is.na(multidrug), 
         g1out_na = is.na(g1out),
         g1hrs_na = is.na(g1hrs), 
         g2hrs_na = is.na(g2hrs))

glm(g1out_na ~ es_g + v_g, 
    data = mis_mech_df, 
    family = "binomial")

glm(multidrug_na ~ es_g + v_g, 
    data = mis_mech_df, 
    family = "binomial")

glm(long_na ~ es_g + v_g, 
    data = mis_mech_df, 
    family = "binomial")

glm(g1hrs_na ~ es_g + v_g, 
    data = mis_mech_df, 
    family = "binomial")

glm(g2hrs_na ~ es_g + v_g, 
    data = mis_mech_df, 
    family = "binomial")
```


```{r}
cor(dd$g1hrs, dd$g2hrs, use = "pairwise.complete")
cor(dd$g1hi, dd$g2hi, use = "pairwise.complete")
```





