---
title: "cca_paper_example"
author: ""
date: "7/6/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages and Data set

```{r, echo=FALSE}
library(ggplot2)
library(metafor)
library(robumeta)
library(metaSEM)
library(naniar)
library(visdat)
data_adt <- readRDS("../data/adt_data.RDS")
```


## Subset of the data 

```{r}
#Subset of variables needed for the analyses:
data_adt_cca<- data_adt[c("studyid", "es_g","se_g","g1hrsperweek","g1txdays")]

#Visualize missingness 
source("../code/wrappers.R")
gg_summary_covariate_miss(data_adt_cca)

data.frame(k.NA=colSums(is.na(data_adt_cca)))
table(rowSums(is.na(data_adt_cca)))

#Percentage missing by variable
miss_var_summary(data_adt_cca)

#Percentage missing by Effect Size case
miss_case_summary(data_adt_cca)

```


## Complete case analysis - Dependent ES

While effect sizes and sampling variances are not missing for this dataset, there are a number of 
moderator (covariates) variables that are. A total of 328 ES available in the data. 

```{r}
##delete NA
data_adt_na<- na.omit(data_adt_cca)

#Meta-regression 
#random-effects model 

#Using rma.mv uncentered group-mean
rma_mv_cca<- rma.mv(es_g, se_g^2, mods= ~ 1 + g1hrsperweek + g1txdays,
                                  random= ~ 1|studyid,
                                  data= data_adt_na,
                                   method="REML")
rma_mv_cca


#Using rma.mv group-mean  centered 
g1hrsperweek.w<- group.center(data_adt_na$g1hrsperweek, grp=data_adt_na$studyid)

rma_mv_cca_2<- rma.mv(es_g, se_g^2, mods= ~ 1 + g1hrsperweek.w + g1txdays,
                                  random= ~ 1|studyid,
                                  data= data_adt_na,
                                   method="REML")
rma_mv_cca_2

```

Only k=247 ES were included in the analysis.Thus, aproximetly 25% of the ES were excluded from the analysis. illustrating the loss of data that can occur when conducting a complete case analysis. 


## Shifting units analysis - Dependent ES

```{r}
#Covariate 1: g1hrsperweek
##delete NA
data_subset<- data_adt_cca[, 1:4]
data_adt_na<- na.omit(data_subset)

#Meta-regression 
#random-effects model 
#Using rma.mv
rma_shu_cov1<- rma.mv(es_g, se_g^2, mods= ~ 1 +  g1hrsperweek,
                                   random= ~ 1|studyid,
                                  data= data_adt_na,
                                  method="REML")
rma_shu_cov1


#Using rma.mv group-mean  centered 
g1hrsperweek.w<- group.center(data_adt_na$g1hrsperweek, grp=data_adt_na$studyid)

rma_shu_cov1<- rma.mv(es_g, se_g^2, mods= ~ 1 +  g1hrsperweek.w,
                                   random= ~ 1|studyid,
                                  data= data_adt_na,
                                  method="REML")
rma_shu_cov1


#Covariate 2: g1txdays
##delete NA
data_subset<- data_adt_cca[,c(1:3, 5)]
data_adt_na<- na.omit(data_subset)

#Meta-regression 
#random-effects model 
#Using rma.mv
rma_shu_cov2<- rma.mv(es_g, se_g^2, mods= ~ 1 +  g1txdays,
                                   random= ~ 1|studyid,
                                  data= data_adt_na,
                                  method="REML")
rma_shu_cov2

```
