---
title: "cca_paper_example"
author: ""
date: "7/6/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages and Data set

```{r, echo=FALSE}
library(ggplot2)
library(metafor)
library(robumeta)
library(metaSEM)
library(naniar)
library(visdat)
data_adt <- readRDS("../data/adt_data.RDS")
```


## Subset of the data 

```{r}
#Subset of variables needed for the analyses:
data_adt_cca<- data_adt[c("studyid", "es_g","se_g", "g1permale", "g1perwhite",  "g1age",
                         "g1hrsperweek","g1txdays")]

#Visualize missingness 
source("../code/wrappers.R")
gg_summary_covariate_miss(data_adt_cca)

data.frame(k.NA=colSums(is.na(data_adt_cca)))
table(rowSums(is.na(data_adt_cca)))

#Percentage missing by variable
miss_var_summary(data_adt_cca)

#Percentage missing by Effect Size case
miss_case_summary(data_adt_cca)

```


## Complete case analysis - Inependent ES

While effect sizes and sampling variances are not missing for this dataset, there are a number of 
moderator (covariates) variables that are. A total of 328 ES available in the data. 

```{r}
#Meta-regression 
#random-effects model 
mr_cca<- rma(es_g, se_g^2, mods= ~ g1permale + g1perwhite + g1age +
                           g1hrsperweek + g1txdays , data= data_adt_cca)
mr_cca

```


Only k=232 ES were included in the analysis.Thus, aproximetly 30% of the ES were excluded from the analysis. illustrating the loss of data that can occur when conducting a complete case analysis. 


# Complete case analysis - Dependet ES (Nested models)

```{r}
##delete NA
data_adt_na<- na.omit(data_adt_cca)

#Meta-regression 
#random-effects model 

#Using rma.mv
rma_mv_cca<- rma.mv(es_g, se_g^2, mods= ~ 1 + g1permale + g1perwhite + g1age +
                                   g1hrsperweek + g1txdays,
                                   random= ~ 1|studyid,
                                  data= data_adt_na,
                                   method="REML")
rma_mv_cca

# Then use robust.rma.mv
robustrma_cca<- robust.rma.mv(rma_mv_cca, data_adt_na$studyid)
robustrma_cca


#Using mr3 with moderators
mr3<- meta3(es_g, se_g^2, cluster= studyid,  data= data_adt_na,
            intercept.constraints = 0,
            x=model.matrix( ~ 1 + g1permale + g1perwhite + g1age +
                           g1hrsperweek + g1txdays))
summary(mr3)

#Using robu
#Intercept only model
#use .8 as correlation between ES
mr_robu<- robu( es_g ~ 1 , data=data_adt_na, studyid, se_g^2, rho=.8, small=TRUE)
print(mr_robu)

sensitivity(mr_robu)

#with moderators
mr_robu2<- robu(es_g ~ g1permale + g1perwhite + g1age +g1hrsperweek + g1txdays ,
                data=data_adt_na, modelweights="HIER", studyid, se_g^2,  small=TRUE)
print(mr_robu2)

```



## Shifting units analysis - Independent ES

```{r}

#Covariate 1: g1permale
mr_shu_cov1<- rma(es_g, se_g^2, mods= ~ g1permale, data= data_adt_cca)
mr_shu_cov1

#319 EF out of 328

#Covariate 2: g1perwhite
mr_shu_cov2<- rma(es_g, se_g^2, mods= ~ g1perwhite, data= data_adt_cca)
mr_shu_cov2

#317 EF out of 328

#Covariate 3: g1age
mr_shu_cov3<- rma(es_g, se_g^2, mods= ~ g1age, data= data_adt_cca)
mr_shu_cov3

#321 EF out of 328

#Covariate 4: g1hrsperweek
mr_shu_cov4<- rma(es_g, se_g^2, mods= ~ g1hrsperweek, data= data_adt_cca)
mr_shu_cov4

#249 EF out of 328

#Covariate 5: g1txdays
mr_shu_cov5<- rma(es_g, se_g^2, mods= ~ g1txdays, data= data_adt_cca)
mr_shu_cov5

#326 EF out of 328

```



## Shifting units of analysis - Dependent ES 

```{r}

#Covariate 1: g1permale
##delete NA
data_subset<- data_adt_cca[, 1:4]
data_adt_na<- na.omit(data_subset)

#Meta-regression 
#random-effects model 
#Using rma.mv
rma_shu_cov1<- rma.mv(es_g, se_g^2, mods= ~ 1 + g1permale,
                                   random= ~ 1|studyid,
                                  data= data_adt_na,
                                  method="REML")
rma_shu_cov1

#319 EF out of 328


#Covariate 2: g1perwhite
##delete NA
data_subset<- data_adt_cca[, c(1:3, 5)]
data_adt_na<- na.omit(data_subset)

#Meta-regression 
#random-effects model 
#Using rma.mv
rma_shu_cov2<- rma.mv(es_g, se_g^2, mods= ~ 1 + g1perwhite,
                                   random= ~ 1|studyid,
                                  data= data_adt_na,
                                  method="REML")
rma_shu_cov2


#Covariate 3: g1age
##delete NA
data_subset<- data_adt_cca[, c(1:3, 6)]
data_adt_na<- na.omit(data_subset)

#Meta-regression 
#random-effects model 
#Using rma.mv
rma_shu_cov3<- rma.mv(es_g, se_g^2, mods= ~ 1 + g1age,
                                   random= ~ 1|studyid,
                                  data= data_adt_na,
                                  method="REML")
rma_shu_cov3


#Covariate 4: g1hrsperweek
##delete NA
data_subset<- data_adt_cca[, c(1:3, 7)]
data_adt_na<- na.omit(data_subset)

#Meta-regression 
#random-effects model 
#Using rma.mv
rma_shu_cov4<- rma.mv(es_g, se_g^2, mods= ~ 1 + g1hrsperweek,
                                   random= ~ 1|studyid,
                                  data= data_adt_na,
                                  method="REML")
rma_shu_cov4


#Covariate 5: g1txdays
##delete NA
data_subset<- data_adt_cca[, c(1:3, 8)]
data_adt_na<- na.omit(data_subset)

#Meta-regression 
#random-effects model 
#Using rma.mv
rma_shu_cov5<- rma.mv(es_g, se_g^2, mods= ~ 1 + g1txdays,
                                   random= ~ 1|studyid,
                                  data= data_adt_na,
                                  method="REML")
rma_shu_cov5

```


